---
title: "Chapter 03"
output: html_notebook
---
date: 2017-08-17
moderator: Jonas

# Chapter 3 summary

* __Counting__: The best thing about Bayesian, you don't can simply think empirically by counting things. Everything is counts, even probability distributions. By using this "empirical attack" on the posterior, a difficult problem of integral calculus is transformed into an easy problem of summarising samples. 

* __Bayesian about Hypothesis testing__: The question in Hypothsis testing is, whether the signal from out model about the Hypothesis is real. The biggest impact on this question is not how well you test, but how realistic you choose your hypothesis. Statistical learning is not about fancy testing, but about thinking. 

* __Describe the Posterior__: The highest posterior density interval is the slimmest interval that holds a given fraction of the posterior samples. It is not often reported but it does make a lot of sense. In the end, you need to ask you, what is the easiest way to report your posterior. If it is a weird skewed one, just give the whole distribution. 

* __Point estimates__: As a researcher you normall want to report a good picture of the posterior, and here point estimates are not helpful. But for decision making, point estimates make sense. In this case, the estimate of choice will depend on the loss-function. So what is the penalty/cost for the estimate being to high or too small. 

#### Average likelihood
In particular, we realized that there is still uncertainty among us how to interpret the average likelihood. In simple examples, this is the sum of the paths to the data. How can we interpret this when we are dealing with more complicated scenarios, e.g. more data, continuous parameter space? Important to remember the function of this: it makes the posterior sum to one (i.e. turn it into a probability distribution).

Furthermore, we should discuss again and track our intuition into the basic functionality of Bayesian updating: that our posterior turns into the new prior upon using additional / new *independent* data.  

#### Use features of posterior for CIs
In general, we concluded that chapter three spoke to our intuition as scientists, and was therefore not terribly hard. We discussed that sampling from the posterior as a way to generate confidence intervals is elegant and intuitive (perhaps unlike the frequentists's interpretation of confidence intervals). In particular, it gives a lot of freedom which intervals to present, based upon choosing areas of the posterior (e.g. the highest posterior density interval, HPDI, that finds the endpoints of the narrowest interval of the posterior where 50% of the probability mass (area) are located). Of course, it must be remembered that the point of this is to usefully summarize the posterior and that of course the entire posterior is relevant but may be unfeasible to show. 

#### Loss
A principled way to not choose the entire posterior is the introduction of a loss function (there are different ones, potentially implying different point estimates). The presented loss function penalizes a parameter choice proportional to the absolute distance of the choice, /disfavoring extreme parameter choices/ as |0.5 - x| for x in [0, 1] is minimized. We then weight our posterior distribution by this loss function to yield the more conservative point estimate of a parameter (/sounds like regularization/?). For example, if you have a posterior that linearly decreases in the interval [0, 1], then the MAP (maximum a posteriori) value would be p = 0. Including loss then favors a more conservative value > 0 (Figure).  
```{r, echo=FALSE}
p_grid <- seq( from=0 , to=1 , length.out=1000 ) 
prior <- rep(1,1000) 
likelihood <- dbinom( 3 , size=3 , prob=p_grid ) 
posterior <- likelihood * prior 
posterior <- posterior / sum(posterior) 
samples <- sample( p_grid , size=1e4 , replace=TRUE , prob=posterior )

loss <- sapply( p_grid , function(d) sum( abs( d - p_grid )))

losspost <- sapply( p_grid , function(d) sum( posterior*abs( d - p_grid )))
#plot
plot(p_grid,loss, type='l', lty=3,lwd=2)
par(new = TRUE)
plot(p_grid,posterior, type = "l", lty=5,lwd=4, axes = FALSE, bty = "n", xlab = "", ylab = "")
par(new = TRUE)
plot(p_grid,losspost, type = "l", pch='O',lwd=4, axes = FALSE, bty = "n", xlab = "", ylab = "")
text(0.5, 0.6, '.. loss\n-- posterior\n- loss*posterior', adj=0)
```

#### Making predictions: propagate uncertainty
The utility and importance of the entire posterior distribution is made clear when discussion prediction---the very point of most modeling. We have two sources of uncertainty: regarding the data, and regarding our parameter choice from fitting. If we chose only a point estimate for our parameter, e.g. the MAP value, our 'prediction' would be 'narrower' and thus overconfident. Instead, we should predict based upon the entire posterior. For this, we choose the parameter values from the entire possible range, and weight /their individual prediction/ by the posterior value for that value. *I wonder if this logic may help rationalize the idea of the 'average likelihood'*?

#### Conclusions and memorable quotes
Both Rene and I really liked some general advice in this chaper (this appears to become a good pattern in this book!).

- *"fetishizing precision to the 5th decimal place will not improve your science"*
- *Don't try and find out __if__ but __how__ your model fails to describe the data. All models are wrong in some respect, so use your judgement.*
- On looking at your data in different ways (e.g. longest sequence of 'Water' outcomes'): *"In your own modeling, you’ll have to imagine aspects of the data that are relevant in your context, for your purposes".* This is key advice to data analysis; and arguably the hardest part.




# Chapter 3 - Exercises

```{r Make data}
p_grid <- seq( from=0 , to=1 , length.out=1000 ) 
prior <- rep( 1 , 1000 ) 
likelihood <- dbinom( 6 , size=9 , prob=p_grid ) 
posterior <- likelihood * prior 
posterior <- posterior / sum(posterior) 
set.seed(100) 
samples <- sample( p_grid , prob=posterior , size=1e4 , replace=TRUE )
```


```{r Easy}
# 3E1. How much posterior probability lies below p = 0.2?
library(tidyverse)
library(rethinking)
dens( samples ) # visual inspection
abline( v=6/9 )
mean( samples<0.2 ) # 0.0005, so very unlikely

# 3E2. How much posterior probability lies above p = 0.8?
mean( samples>0.8 ) # 0.1117, so much more

# 3E3. How much posterior probability lies between p = 0.2 and p = 0.8?
mean( samples>0.2 & samples<0.8 ) # 0.8878

# 3E4. 20% of the posterior probability lies below which value of p?
samples %>% quantile( 0.2 ) # 0.5195195

# 3E5. 20% of the posterior probability lies above which value of p?
samples %>% quantile( 1-0.2 ) # 0.7568

# 3E6. Which values of p contain the narrowest interval equal to 66% of the posterior probability?
HPDI( samples , prob=0.66 ) # 0.52 0.787
dens( samples )
abline( v=0.52 ); abline( v=0.784 )

# 3E7. Which values of p contain 66% of the posterior probability, assuming equal posterior probability both below and above the interval?
samples %>% quantile( c(1/6, 5/6) )
dens( samples )
abline( v=0.4985 ); abline( v=0.77 )
```
```{r Medium}
## Medium.

# 3M1. Suppose the globe tossing data had turned out to be 8 water in 15 tosses. Construct the posterior distribution, using grid approximation. Use the same flat prior as before.

p_grid <- seq( from=0 , to=1 , length.out=1000 )
prior <- rep( 1, length( p_grid ) )
likelihood <- dbinom( x=8 , size=15 , prob=p_grid )
posterior <- prior * likelihood
posterior <- posterior / sum( posterior )
plot( p_grid , posterior , type ='l' )

# 3M2. Draw 10,000 samples from the grid approximation from above. Then use the samples to calculate the 90% HPDI for p.
samples <- sample( p_grid , prob=posterior , size=1e4 , replace=T )
dens( samples )
HPDI( samples , prob=.90 )

# 3M3. Construct a posterior predictive check for this model and data. This means simulate the distribution of samples, averaging over the posterior uncertainty in p. What is the probability of observing 8 water in 15 tosses?
ppc <- rbinom( n=length(samples) , size=15 , prob=samples )  
simplehist( ppc )
mean( ppc == 8 ) # it is 0.1397

# 3M4. Using the posterior distribution constructed from the new (8/15) data, now calculate the probability of observing 6 water in 9 tosses.
# I guess this means using the posterior now as a prior
prior2 <- posterior
likelihood <- dbinom( x=6 , size=9 , prob=p_grid )
posterior2 <- prior2 * likelihood
posterior2 <- posterior2 / sum( posterior2 )
plot( p_grid , posterior2 , type='l' )

# 3M5. Start over at 3M1, but now use a prior that is zero below p = 0.5 and a constant above p = 0.5. This corresponds to prior information that a majority of the Earth’s surface is water. Repeat each problem above and compare the inferences. What difference does the better prior make? If it helps, compare inferences (using both priors) to the true value p = 0.7.
priorM5 <- ifelse( p_grid<0.5 , 0 , 1 )
#
likelihood <- dbinom( x=8 , size=15 , prob=p_grid )
# old
posterior <- prior * likelihood
posterior <- posterior / sum( posterior )
plot( p_grid , posterior , type ='l' )
# new
posterior_n <- priorM5 * likelihood
posterior_n <- posterior_n / sum( posterior_n )
plot( p_grid , posterior_n , type ='l' )
# where there is nothing in the prior, there is nothing in the posterior

# old
samples <- sample( p_grid , prob=posterior , size=1e4 , replace=T )
dens( samples )
HPDI( samples , prob=.90 )
# new
samples_n <- sample( p_grid , prob=posterior_n , size=1e4 , replace=T )
dens( samples_n )
HPDI( samples_n , prob=.90 )

# old
ppc <- rbinom( n=length(samples) , size=15 , prob=samples )  
simplehist( ppc )
mean( ppc == 8 ) # it is 0.1504
# new
ppc_n <- rbinom( n=length(samples_n) , size=15 , prob=samples_n )  
simplehist( ppc_n )
mean( ppc_n == 8 ) # it is 0.159, is higher than with other prior

# old
prior2 <- posterior
likelihood <- dbinom( x=6 , size=9 , prob=p_grid )
posterior2 <- prior2 * likelihood
posterior2 <- posterior2 / sum( posterior2 )
plot( p_grid , posterior2 , type='l' )
# new
prior2 <- posterior_n
likelihood <- dbinom( x=6 , size=9 , prob=p_grid )
posterior2 <- prior2 * likelihood
posterior2 <- posterior2 / sum( posterior2 )
plot( p_grid , posterior2 , type='l' )

# with the new prior, there is never any prob in the posterior for p<0.5. 
# the predictive checks are more consistant with the data. 
```

```{r Hard}
## Hard

# data
birth1 <- c(1,0,0,0,1,1,0,1,0,1,0,0,1,1,0,1,1,0,0,0,1,0,0,0,1,0, 0,0,0,1,1,1,0,1,0,1,1,1,0,1,0,1,1,0,1,0,0,1,1,0,1,0,0,0,0,0,0,0, 1,1,0,1,0,0,1,0,0,0,1,0,0,1,1,1,1,0,1,0,1,1,1,1,1,0,0,1,0,1,1,0, 1,0,1,1,1,0,1,1,1,1) # first born children
birth2 <- c(0,1,0,1,0,1,1,1,0,0,1,1,1,1,1,0,0,1,1,1,0,0,1,1,1,0, 1,1,1,0,1,1,1,0,1,0,0,1,1,1,1,0,0,1,0,1,1,1,1,1,1,1,1,1,1,1,1,1, 1,1,1,0,1,1,0,1,1,0,1,1,1,0,0,0,0,0,0,1,0,0,0,1,1,0,0,1,0,0,1,1, 0,0,0,1,1,1,0,0,0,0) # second born children 

data(homeworkch3)
birth2

# all boys born as first-born
sum( birth1 )
# all boys born as second born
sum( birth2 )

# 3H1. Using grid approximation, compute the posterior distribution for the probability of a birth being a boy. Assume a uniform prior probability. 

p_vec <- seq( from=0 , to=1 , length.out=1000 ) # parameter p and its conjectures
prior <- rep( 1 , length( p_vec ) ) # boy , girl, [counts of these 2 conjectures to happen, without data]
## turn data into observation
all_births <- c( birth1 , birth2 )
nboys <- sum( all_births ) # 111
size <- length( all_births ) # 200
##
likelihood <- dbinom( x=nboys , size=size , prob=p_vec )
# update prior
posterior <- prior * likelihood
posterior <- posterior / sum( posterior )
plot( p_vec , posterior , type='l' )

# Which parameter value maximizes the posterior probability?
p_vec[ which.max(posterior) ] # 0.554

# 3H2. Using the sample function, draw 10,000 random parameter values from the posterior distribution you calculated above. Use these samples to estimate the 50%, 89%, and 97% highest posterior density intervals.
samples <- sample( p_vec , prob=posterior , size=1e4 , replace=T )
samples %>% HPDI(0.5)
samples %>% HPDI(0.89)
samples %>% HPDI(0.97)

# 3H3. Use rbinom to simulate 10,000 replicates of 200 births. You should end up with 10,000 numbers, each one a count of boys out of 200 births. Compare the distribution of predicted numbers of boys to the actual count in the data (111 boys out of 200 births). There are many good ways to visualize the simulations, but the dens command (part of the rethinking package) is probably the easiest way in this case. Does it look like the model fits the data well? That is, does the distribution of predictions include the actual observation as a central, likely outcome?

fp_sims <- rbinom( n=1e4 , size=200 , prob=samples )
fp_sims %>% length()
dens( fp_sims ); abline( v=111 ) # it looks like a good fit indeed

# 3H4. Now compare 10,000 counts of boys from 100 simulated first borns only to the number of boys in the first births, birth1. How does the model look in this light?

# now different data
size <- length( birth1 )
nboys <- sum( birth1 )
likelihood <- dbinom( x=nboys , size=size , prob=p_vec )
posterior <- prior * likelihood
posterior <- posterior / sum( posterior )
plot( p_vec , posterior , type='l' )

# posterior samples
samples <- sample( p_vec , prob=posterior , size=1e4 , replace=T )
fp_sims <- rbinom( n=1e4 , size=100 , samples )
dens( fp_sims ); abline( v=51 , col='red' )
simplehist( fp_sims ); abline( v=51 , col='red')

# 3H5. The model assumes that sex of first and second births are independent. To check this assumption, focus now on second births that followed female first borns. Compare 10,000 simulated counts of boys to only those second births that followed girls. To do this correctly, you need to count the number of first borns who were girls and simulate that many births, 10,000 times. Compare the counts of boys in your simulations to the actual observed count of boys following girls. How does the model look in this light? Any guesses what is going on in these data?

b2_aftergirl <- birth2[birth1 == 0]
size <- length( b2_aftergirl ) # 49
nboys <- sum( b2_aftergirl ) # 39
likelihood <- dbinom( x=nboys , size=size , prob=p_vec )
posterior <- prior * likelihood
posterior <- posterior / sum( posterior )
plot( p_vec , posterior , type='l' )

# posterior sample and future prodiction simulation
samples <- sample( p_vec , prob=posterior , size=1e4 , replace=T )
fp_sims <- rbinom( n=1e4 , size=size , prob=samples )
dens( fp_sims/size , adj=0.9); abline( v=39/size ,col='red') # it still looks good

# What is going on in the data?
# The data seems to support a model, where after having a girl it is much more likely having a girl. This could be because families, after having one child of each sex stopped having babies, and there are therefore concentrated in the 2-child group of families. 
```




