---
title: "jonas_chapter_3"
output: html_notebook
---

# Sampling from the posterior is the name of the game

```{r}
p_grid <- seq( from=0 , to=1 , length.out=1000 ) 
prior <- rep( 1 , 1000 ) 
likelihood <- dbinom( 6 , size=9 , prob=p_grid ) 
posterior <- likelihood * prior 
posterior <- posterior / sum(posterior)
# sample from posterior
samples <- sample( p_grid , prob=posterior , size=1e4 , replace=TRUE )
plot(samples)
```
```{r}
library(rethinking)
dens( samples )
```
```{r}
# add up posterior probability where p < 0.5 
print(sum( posterior[ p_grid < 0.5 ] ))
# instead, find the frequency of parameter values below 0.5 for a method that works with models of more parameters
print(sum( samples < 0.5) /1e4 )

# or for another interval
print( sum( samples > 0.5 & samples < 0.75) / 1e4)

# or to find the boundaries of the lower 80% of posterior probability
print(quantile( samples , 0.8 ))

# or to find the middle 80% (i.e. the middle 80% interval lies between the 10th percentile and the 90th percentile.)

print(quantile( samples , c( 0.1 , 0.9 ) ))
```
In terms of supporting inferences about swhich parameters are consisten with the data, these intervals are not perfect (especially when the posterior is funnily skewed.)


```{r}
p_grid <- seq( from=0 , to=1 , length.out=1000 ) 
prior <- rep(1,1000) 
likelihood <- dbinom( 3 , size=3 , prob=p_grid ) 
posterior <- likelihood * prior 
posterior <- posterior / sum(posterior) 
samples <- sample( p_grid , size=1e4 , replace=TRUE , prob=posterior )

# the 50% percentile confidence interval is computed by the PI function
PI( samples, prob = 0.5)

# choose the 'highest posterior density interval' - the narrowest interval contraining the specified probability mass.
HPDI( samples, prob = 0.5)
```
To get a point estimate:

```{r}
print(p_grid[ which.max(posterior) ])

print(chainmode( samples , adj=0.01 ))

```


Understanding loss: absolute distance.
Minimal at 0.5 in [0,1].
Maximal at 1 or 0 in [0, 1]
```{r}
print('loss w/o posterior info')
print( sum(abs(0.5 - p_grid)))
print( sum(abs(0.0 - p_grid)))
print( sum(abs(1.0 - p_grid)))
png("./loss.png",)

loss <- sapply( p_grid , function(d) sum( abs( d - p_grid )))

print('loss with posterior')
print( sum(posterior*abs(0.5 - p_grid)))
print( sum(posterior*abs(0.0 - p_grid)))
print( sum(posterior*abs(1.0 - p_grid)))
losspost <- sapply( p_grid , function(d) sum( posterior*abs( d - p_grid )))
#plot
plot(p_grid,loss, type='l', lty=3,lwd=2)
par(new = TRUE)
plot(p_grid,posterior, type = "l", lty=5,lwd=4, axes = FALSE, bty = "n", xlab = "", ylab = "")
par(new = TRUE)
plot(p_grid,losspost, type = "l", pch='O',lwd=4, axes = FALSE, bty = "n", xlab = "", ylab = "")
text(0.5, 0.6, '.. loss\n-- posterior\n- loss*posterior', adj=0)
dev.off()
```

USing a loss function as a principled way to figure out a point estimate

```{r}
expected_loss_guess_0_5 = sum(posterior * abs(0.5 - p_grid))
print(expected_loss_guess_0_5)

#all loss (repeat above for each value)
loss <- sapply( p_grid , function(d) sum( posterior*abs( d - p_grid ) ))
plot(p_grid,loss)

```

```{r}

dummy_v <- rbinom(1e5, size=1e3, prob=0.7)
simplehist( dummy_v, xlab="dummy water count")
```

# Dummy data
```{r}
dbinom( 0:2 , size=2 , prob=0.7 )
hist(rbinom( 500 , size=2 , prob=0.7 ))
```














